---
title: "Assignment 2"
author: "Erik Nylander"
date: "Tuesday, February 03, 2015"
output: html_document
---

### Problem Set 1

1) Show that $A^T A \neq A A^T$ in general.
Lets assume that $A^T A = A A^T$ and see if there is a counter example.
Let A be the matrix: $A = \begin{bmatrix} a & b & c \\ d & e & f \end{bmatrix}$ and 
$A^T = \begin{bmatrix} a & d \\ b & e \\ c & f \end{bmatrix}$ therefore:
$$
A^T A = 
\begin{bmatrix} a & d \\ b & e \\ c & f \end{bmatrix}
\begin{bmatrix} a & b & c \\ d & e & f \end{bmatrix} =
\begin{bmatrix} 
a^2 + d^2 & ab + de & ac + df \\ 
ba + ed & b^2 + e^2 & bc + ef \\
ca+fd & cb + fe & c^2 + f^2 
\end{bmatrix}
$$
$$
A A^T = 
\begin{bmatrix} a & b & c \\ d & e & f \end{bmatrix}
\begin{bmatrix} a & d \\ b & e \\ c & f \end{bmatrix} =
\begin{bmatrix}
a^2 + b^2 + c^2 & ad + be + cf \\
da + eb + fc & d^2 + e^2 + f^2
\end{bmatrix}
$$
Since $A^T A \neq A A^T$ in our example we know that in general $A^T A \neq A A^T$


2) $A^T A = A A^T$ is true if $A$ is a diagonal matrix. 
Let $A = \begin{bmatrix} a & 0 & 0 \\ 0 & b & 0 \\ 0 & 0 & c \end{bmatrix}$ then
$A^T = \begin{bmatrix} a & 0 & 0 \\ 0 & b & 0 \\ 0 & 0 & c \end{bmatrix}$ therefore:
$$
A^T A =
\begin{bmatrix} a & 0 & 0 \\ 0 & b & 0 \\ 0 & 0 & c \end{bmatrix}
\begin{bmatrix} a & 0 & 0 \\ 0 & b & 0 \\ 0 & 0 & c \end{bmatrix} =
\begin{bmatrix} a^2 & 0 & 0 \\ 0 & b^2 & 0 \\ 0 & 0 & c^2 \end{bmatrix}
$$
$$
A A^T =
\begin{bmatrix} a & 0 & 0 \\ 0 & b & 0 \\ 0 & 0 & c \end{bmatrix}
\begin{bmatrix} a & 0 & 0 \\ 0 & b & 0 \\ 0 & 0 & c \end{bmatrix} =
\begin{bmatrix} a^2 & 0 & 0 \\ 0 & b^2 & 0 \\ 0 & 0 & c^2 \end{bmatrix}
$$
This gives a case when $A^T A = A A^T$ is true.

### Problem Set 2

To perform the LU decomposition of the matrix I'll be taking the given matrix, generating an identity matrix and using it to store the steps for deconstructing
the matrix. The function will return the lower and upper matrices in a vector.

```{r}
LUDecomp <- function(U){
    # Computes the LU decomposition of square matrix.
    #
    # Args:
    #   U = A square matrix
    #
    # Returns:
    #   result = A list containg the lower and upper diagonal matrices.
    #
    
    L<-diag(dim(U)[1]) # Identity matrix that stores the elimination matrices
    # Outer for loop traverses the rows and the inner for loop traverses the columns
    for (i in 2:dim(U)[2]){
        for (n in 1:(i-1)){
            E<-diag(dim(U)[2])
            E[i,n]<- -(U[i,n]/U[n,n]) # Generates the Elimination matrix for each step
            U <- E %*% U # Eliminates the element from the upper diagonal matrix
            L <- L %*% solve(E) # Stores the elimination step
        }
    }
    result <- list('lower'=L, 'upper'=U)
    return(result)
}
```

#### Testing with a 2X2 Matrix
```{r}
(A1 <- matrix(c(-3,1,6,-8), nrow=2))
(R1 <- LUDecomp(A1))
R1$lower %*% R1$upper
```

#### Testing with a 3X3 Matrix
```{r}
(A2 <- matrix(c(1,2,3,1,1,1,2,0,1), nrow=3))
(R2 <- LUDecomp(A2))
R2$lower %*% R2$upper
```

#### Testing with a 5X5 Matrix
```{r}
set.seed(42)
(A3 <- matrix(sample(1:10, 25, replace=TRUE), nrow=5))
(R3 <- LUDecomp(A3))
R3$lower %*% R3$upper
```