---
title: "Assignment 1"
author: "Erik Nylander"
date: "Friday, January 30, 2015"
output: html_document
---

### Problem 1

Given $u = \begin{bmatrix} 0.5\\ 0.5 \end{bmatrix}$ and $v = \begin{bmatrix} 3\\ -4 \end{bmatrix}$


1) Calculate the dot product $u \cdot v$.
$$ 
\begin{bmatrix} 0.5\\ 0.5 \end{bmatrix} 
\cdot 
\begin{bmatrix} 3\\ -4 \end{bmatrix} = 
0.5 * 3 + 0.5 * -4 = -0.5
$$

2) What are the lengths of $u$ and $v$?
$$
\| u \| = 
\sqrt{u \cdot u} =
\sqrt{0.5^2 + 0.5^2} =
\sqrt{0.25} =
.707
$$
$$
\| v \| =
\sqrt{v \cdot v} =
\sqrt{3^2 + (-4)^2} =
\sqrt{25} =
5
$$

3) What is the linear combination: $3u - 2v$
$$
3u - 2v = 
3 \begin{bmatrix} 0.5\\ 0.5 \end{bmatrix} -
2 \begin{bmatrix} 3\\ -4 \end{bmatrix} =
\begin{bmatrix} 1.5\\ 1.5 \end{bmatrix} -
\begin{bmatrix} 6\\ -8 \end{bmatrix} =
\begin{bmatrix} -4.5\\ 9.5 \end{bmatrix}
$$

4) What is the angle between$u$ and $v$?
$$
cos x = 
\frac{u \cdot v}{\| u \| \| v \|} =
\frac{-0.5}{.707 * 5}
$$
$$
x = \arccos (\frac{-0.5}{.707 * 5}) = 1.712  98.13^\circ
$$

### Problem 2

To solve the system of equations using 3 variable and 3 constraints I'll be using the elimination method to construct the upper triangular form. The only modification that I will be using is to create and augmented matrix from the coefficient matrix and constraints vector. This allows for easier manipulation of the constraints.

```{r}
solveByElimination <- function(A,b){
    # Solves a 3X3 matrix by elimination
    #
    # Args:
    #   A: A 3X3 matrix of the coefficients
    #   b: A vector of the constraints
    #
    # Returns:
    #   A vector of the solutions to the system of equations.
    
    A = cbind(A,b)
    # Preforms Row swaps if A[1,1] is 0
    if(A[1,1] == 0){
        # Checks if both A[1,1] and A[2,1] are 0 and Swaps
        if(A[2,1] == 0){
           A <- A[c(3,1,2),]
        } else {
           A <- A[c(2,1,3),]
        }
        
    }
    # Uses A[1,1] as the pivot to eliminate the other values in the column
    A[2,] <- A[2,] - ((A[2,1]/A[1,1])*A[1,])
    A[3,] <- A[3,] - ((A[3,1]/A[1,1])*A[1,])
    
    # Checks to see if A[2,2] is a zero and swaps the row if it is.
    if (A[2,2] == 0){
        A <- A[c(1,3,2),]
    }
    # Uses A[2,2] as the pivot to eliminate the value in row 3
    A[3,] <- A[3,] - ((A[3,2]/A[2,2]*A[2,]))
    
    # Solving by back subsitution
    x3 <- A[3,4]/A[3,3]
    x2 <- (A[2,4]-A[2,3]*x3)/A[2,2]
    x1 <- (A[1,4]-A[1,2]*x2-A[1,3]*x3)/A[1,1]
    
    result <- c(x1,x2,x3)
    return (result)
}
```

Now I will check the function using the provided example of the following system:
$$
\begin{bmatrix} 1 & 1 & 3 \\ 2 & -1 & 5 \\ -1 & -2 & 4 \end{bmatrix}
\begin{bmatrix} x_1 \\ x_2 \\ x_3 \end{bmatrix} =
\begin{bmatrix} 1 \\ 2 \\ 6 \end{bmatrix}
$$

```{r}
mat <- matrix(c(1,2,-1,1,-1,-2,3,5,4), ncol=3)
con <- c(1,2,6)
solveByElimination(mat, con)
```

This yields the solution:
$$
x = \begin{bmatrix} -1.545 \\ -0.318 \\ 0.954 \end{bmatrix}
$$